Type Error Telemetry at Scale

### 2023-09-30 programming23/paper.tex
sec 4 pass done, removed timelines

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}
\newcommand{\percentile}[1]{{#1}th~percentile}
\newcommand{\ptile}[1]{\textsc{p}{#1}}
\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}
\newcommand{\FS}{background}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\zeroheight}[1]{\raisebox{0pt}[0pt][0pt]{\(#1\)}}
\newcommand{\nK}[1]{#1\textsc{k}}
\newcommand{\stddev}[1]{\nK{#1}}
\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
\newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1}}
\newcommand{\obox}[1]{\colorbox{orange!29}{#1}}
\newcommand{\rbox}[1]{\colorbox{red!30}{#1}}
\newcommand{\tefsfont}[1]{\textsc{#1}}
\newcommand{\tekey}{\tefsfont{te}}
\newcommand{\fskey}{\tefsfont{fs}}
\newcommand{\QALAN}{\FILL{} Alan:}
\newcommand{\panon}{pseudonymized}
\newcommand{\symbox}[1]{{\makebox[0.8em][c]{{#1}}}}
\newcommand{\addsym}{\symbox{\(\uparrow\)}}
\newcommand{\keepsym}{\symbox{\(=\)}}
\newcommand{\dropsym}{\symbox{\(\downarrow\)}}
% \newcommand{\dsymbox}[1]{\makebox[0.8em][c]{#1}}
% \newcommand{\dropsym}{\dsymbox{\raisebox{0ex}[0.9em]{\raisebox{0.5ex}{\vrule width 0.6em height 0.6pt}}}}

@@

Our starting point is a telemetry framework that is built in
to {Roblox Studio} and currently measures the effectiveness of
creation features.
This system randomly determines which sessions should report telemetry, and,
for those sessions, reports telemetry records back with a summary of the
session.

@@

\section{Context: {Roblox} and {Luau}}
\label{s:context}
% https://create.roblox.com/docs/scripting/luau

% TODO quick reference table
% \begin{table}
%   \caption{Modes and analyses}
%   \label{t:mode-analyze}
% 
%   Mode: nocheck ...
%   Analysis: nocheck ... forced strict
% 
% \end{table}

@@

  % TE: nocheck & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
  % TE: nonstrict & 19491 & [\pct{39.63}] & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
  % TE: strict & 733 & [\pct{39.18}] & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
  %
  % TE mod: nocheck & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
  % TE mod: nonstrict & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
  % TE mod: strict & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
  %
  % FS: nocheck & 270763 & [\pct{36.99}] & 238696 & [\pct{32.61}] & 222588 & [\pct{30.41}] \\
  % FS: nonstrict & 35330 & [\pct{37.28}] & 29483 & [\pct{31.11}] & 29955 & [\pct{31.61}] \\
  % FS: strict & 574 & [\pct{33.71}] & 561 & [\pct{32.94}] & 568 & [\pct{33.35}] \\
  %
  % FS mod: nocheck & 259865 & [\pct{39.09}] & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
  % FS mod: nonstrict & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
  % FS mod: strict & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
  \begin{tabular}{l@{~~}lr@{}rr@{}rr@{}r}
    & & \zerowidth{\rbox{\addsym{}}} & & \zerowidth{\ybox{\keepsym{}}} & & \zerowidth{\gbox{\dropsym{}}} \\\midrule
    type      & \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
    error     & \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
              & \mstrict{}    & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
    \\
    \FS       & \mnocheck{}   & 259865 & [\pct{39.09}] & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
    error     & \mnonstrict{} & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
              & \mstrict{}    & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
  \end{tabular}
  \caption{How often do edits increase, maintain, or decrease the number of errors?}
  \label{f:error-changes}
\end{figure}

@@

  \item
    \mstrict{} mode has a higher percent of  (\keepsym{}) records.
    It may be that errors are more likely to persist in this mode.
    %% TODO check if actually zero

@@

\begin{figure}[t]\centering
  %% TODO
  %% - thin bars
  %% - colors for good , bad (disagree) , neutral
  \begin{tabular}{lll}
    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
  \end{tabular}
  \caption{Comparing the change in type errors to the change in \FS{} errors.}
  \label{f:tefs-compass}
\end{figure}

Since Roblox Studio runs each codebase through two passes of type analysis,
it is interesting to compare their results.
Of special interest is the \mstrict{} mode case, because the number of
type errors and \FS{} errors should be nearly the same.
But, there are two differences between \mstrict{} type checking
and \FS{} checking:
\begin{itemize}
  \item
    Background checks all dependencies in \mstrict{} mode, no matter
    what mode they declare.
    So, the number of \FS{} errors may be higher.
  \item
    Background checking has looser rules for the \emph{data model};
    that is, for the static assets that a program manipulates.
    So, the number of \FS{} errors may be lower.
    %% initial DM, all properties go into a table type; DM -> folder -> model -> part
    %% get dot-driven autocomplete
\end{itemize}
Our data contains a surprising number of records where \FS{}
reports fewer errors.
Data model edits are apparently a regular occurrence in Luau programs.

\Cref{f:tefs-compass} compares changes in type errors (\tekey{}) to changes in
\FS{} errors (\fskey{}) for the three analysis modes.
To a first approximation, the changes either agree with one another
or disagree.
For example, if type errors increase but \FS{} errors decrease,
we have a disagreement.
The plots futher categorize agreements and disagreements into
the following five cases:
\begin{enumerate}
  \item
    \emph{zero}: neither \tekey{} nor \fskey{} changed
  \item
    \emph{te only}: \tekey{} changed but \fskey{} remained the same
  \item 
    \emph{fs only}: \tekey{} remained the same but \fskey{} changed
  \item
    \emph{dis}: \tekey{} and \fskey{} changed in opposite directions
  \item 
    \emph{agree}: \tekey{} and \fskey{} changed in the same direction
\end{enumerate}
Both \emph{zero} and \emph{agree} are agreements.
The others are disagreements.
The plots show what percent of all telemetry records fell into these five categories.

@@

\begin{itemize}
  \item
    All three plots have a nonzero percentage of \emph{te only} items.
    In \mstrict{} mode, the percentage jumps up to nearly \pct{25}.
    Since \fskey{} is more lenient only for data model changes, these must
    be edits that affect the model in a way that raises a type error.
    Naturally, \mstrict{} mode has the largest number of such edits because
    its type checker is the most careful.

  \item
    In the first plot, for \mnocheck{},
    half the records are \emph{zero\/}s, nearly half are \emph{fs only}, and only a few
    land in the other categories.
    While the number of zeros is a bit high,
    the \emph{fs. only} spike makes sense because type checking is supposed to complain
    sparingly in \mnocheck{} mode.

  \item
    In the \mnonstrict{} plot, the number of agreements jumps up.
    This is reassuring.
    \FILL{}.

@@

Users of \mstrict{} mode may find it frustrating that the type checker complains
about data model edits.
Though we cannot see the exact reason for the complaints, the checker errs on
the conservative side.
Thus we expect the number of useful reports is low.
This may be a barrier to adaption.

\subsection{Session Timelines}

\begin{figure}[t]\centering
  %% TODO any way to cluster timelines to avoid the "creepiness" of spying on one user?
%  \mnonstrict{}\\
%  \includegraphics[width=\columnwidth]{img/timeline-nonstrict.pdf}

  \medskip{}
  \mstrict{}\\
  \includegraphics[width=\columnwidth]{img/timeline-strict.pdf}

  \caption{Two example sessions.}
  \label{f:indy-session}
\end{figure}

Another insightful visualization that the telemetry supports is plots of individual sessions
over time.
\Cref{f:indy-session} shows the type errors, \FS{} errors, and module switches
in two \mstrict{} mode sessions.
The x-axis shows time in seconds from the start of the session.
The y-axis counts errors.
Type errors appear as orange plus marks.
Background errors appear as green x marks.
The vertical lines show module switches.

In the first plot, the error counts are low until about the 3,000s mark,
at which point an edit (not a module switch) leads to 15 errors.
These 15 errors disappear two records later.
Then, after two module switches and some edits, the number of errors spikes again.
In the second plot, the number of errors spikes in the beginning due to edits,
then remains low for the next few thousand seconds.

A general observation is that the number of type errors is always less than the number of \FS{}
errors, though the numbers are close.
This should happen in \mstrict{} mode.

@@

On the meta level, the main takeaway is that the lightweight, counter-based telemetry
in Roblox Studio supports a variety of useful analyses:
\begin{itemize}
  \item
    Counting specific type errors helps to identify trends and to confirm that
    undesirable events rarely happen (e.g.,\code{UnificationTooComplex}).
  \item
    Testing the relationship between two counts can yield insights.
    The gap between \mstrict{} and \FS{} errors in~\cref{f:tefs-compass}
    revealed an issue with \mstrict{} checking.
  \item
    Reporting the overlap between type errors and the current (approximate) edit
    range lets us assess the outcome of edits without revealing source code.
    It was especially useful to have overlaps for two sets of type errors: old and current.
  \item
    Timestamps give useful, low-level insights about the frequency of keystrokes,
    the length of sessions, and actions over time~(\cref{f:indy-session}).
    The event ordering that timestamps provide was invaluable; e.g., for learning
    that error deltas bounce up and down rather than rising shaply, then falling~(\cref{f:error-density}).
\end{itemize}

>>>>

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}
\newcommand{\percentile}[1]{{#1}th~percentile}
\newcommand{\ptile}[1]{\textsc{p}{#1}}
\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}
\newcommand{\FS}{background}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\zeroheight}[1]{\raisebox{0pt}[0pt][0pt]{\(#1\)}}
\newcommand{\nK}[1]{#1\textsc{k}}
\newcommand{\stddev}[1]{\nK{#1}}
\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
\newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1}}
\newcommand{\obox}[1]{\colorbox{orange!29}{#1}}
\newcommand{\rbox}[1]{\colorbox{red!30}{#1}}
\newcommand{\tefsfont}[1]{\textsc{#1}}
\newcommand{\tekey}{\tefsfont{te}}
\newcommand{\fskey}{\tefsfont{bg}}
\newcommand{\QALAN}{\FILL{} Alan:}
\newcommand{\panon}{pseudonymized}
\newcommand{\symbox}[1]{{\makebox[0.8em][c]{{#1}}}}
\newcommand{\addsym}{\symbox{\(\uparrow\)}}
\newcommand{\keepsym}{\symbox{\(=\)}}
\newcommand{\dropsym}{\symbox{\(\downarrow\)}}
% \newcommand{\dsymbox}[1]{\makebox[0.8em][c]{#1}}
% \newcommand{\dropsym}{\dsymbox{\raisebox{0ex}[0.9em]{\raisebox{0.5ex}{\vrule width 0.6em height 0.6pt}}}}

@@

Our starting point is a telemetry framework that is built in
to {Roblox Studio} and currently measures the effectiveness of
creation features.
%% \QALAN: can we be more specific? what's an example ... API usage?
This system randomly determines which sessions should report telemetry, and,
for those sessions, reports telemetry records back with a summary of the
session.

@@

\section{Context: {Roblox} and {Luau}}
\label{s:context}

% https://create.roblox.com/docs/scripting/luau

@@

  % TE: nocheck & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
  % TE: nonstrict & 19491 & [\pct{39.63}] & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
  % TE: strict & 733 & [\pct{39.18}] & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
  %
  % TE mod: nocheck & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
  % TE mod: nonstrict & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
  % TE mod: strict & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
  %
  % FS: nocheck & 270763 & [\pct{36.99}] & 238696 & [\pct{32.61}] & 222588 & [\pct{30.41}] \\
  % FS: nonstrict & 35330 & [\pct{37.28}] & 29483 & [\pct{31.11}] & 29955 & [\pct{31.61}] \\
  % FS: strict & 574 & [\pct{33.71}] & 561 & [\pct{32.94}] & 568 & [\pct{33.35}] \\
  %
  % FS mod: nocheck & 259865 & [\pct{39.09}] & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
  % FS mod: nonstrict & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
  % FS mod: strict & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
  \begin{tabular}{l@{~~}lr@{}rr@{}rr@{}r}
    & & \zerowidth{\rbox{\addsym{}}} & & \zerowidth{\ybox{\keepsym{}}} & & \zerowidth{\gbox{\dropsym{}}} \\\midrule
    type      & \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
    error     & \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
              & \mstrict{}    & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
    \\[-2ex]
    \FS       & \mnocheck{}   & 259865 & [\pct{39.09}] & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
    error     & \mnonstrict{} & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
              & \mstrict{}    & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
  \end{tabular}
  \caption{How often do edits increase, maintain, or decrease the number of errors?}
  \label{f:error-changes}
\end{figure}

@@

  \item
    \mstrict{} mode has a higher percent of ``maintain'' (\keepsym{}) records.
    Errors may be more likely to persist in this mode.
    %% TODO check if actually zero

  \item
    In \FS{} mode, the percent of ``maintain'' records is high across
    the board and the percents of increases and decreases are lower.
    No matter the mode, reducing type errors is not guaranteed to reduce
    \FS{} errors as well.
    The next section further compares type and \FS{} errors.

@@

\begin{figure}[t]\centering
  %% TODO
  %% - thin bars
  %% - colors for good , bad (disagree) , neutral
  \begin{tabular}{lll}
    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
    \includegraphics[width=0.26\columnwidth]{img/compass-nocheck.pdf}
    &
    \includegraphics[width=0.26\columnwidth]{img/compass-nonstrict.pdf}
    &
    \includegraphics[width=0.26\columnwidth]{img/compass-strict.pdf}
  \end{tabular}
  \caption{Comparing the change in type errors to the change in \FS{} errors.}
  \label{f:tefs-compass}
\end{figure}

\Cref{f:tefs-compass} takes a closer look at type errors (\tekey{}) and \FS{} errors (\fskey{}).
For every pair of events in each session, we compare the change in
type errors to the change in \FS{} errors and choose one of five categories.
At a high level, the question at hand is whether the changes \emph{agree} or \emph{disagree}.
The categories go one step further and separate zero and non-zero changes:

\noindent{}\begin{tabular}{lll}
  1. & \code{zero} &  neither \tekey{} nor \fskey{} changed \\
  2. & \code{te only} & \tekey{} changed but \fskey{} remained the same \\
  3. & \code{bg only} & \tekey{} remained the same but \fskey{} changed \\
  4. & \code{dis} & \tekey{} and \fskey{} changed in opposite directions \\
  5. & \code{agree} & \tekey{} and \fskey{} changed in the same direction
\end{tabular}
\smallskip{}

Of particular interest is how \mstrict{} and \FS{} checks compares.
The two analyses are similar, but \FS{} analysis uses strong checks
on all dependencies and has looser rules for the data model~(\cref{s:context}):
whereas \mstrict{} requires casts to confirm the types of data assets,
\FS{} gives data the dynamic gradual type.

@@

\begin{itemize}
  \item
    All three plots have a nonzero percentage of \code{te only} items.
    In \mstrict{} mode, the percentage jumps up to nearly \pct{25}.
    Since \FS{} analysis is more lenient only for data model changes, these must
    be edits that affect the model in a way that raises a type error.
    The high percentage in \mstrict{} may be a source of frustration.
    %% Naturally, \mstrict{} mode has the largest number of such edits because
    %% its type checker is the most careful.

  \item
    In the first plot, for \mnocheck{},
    half the records are \code{zero}, nearly half are \code{bg only}, and only a few
    land in the other categories.
    While the number of zeros is a bit high, the high \code{bg only} and low
    \code{agree} counts make sense because \mnocheck{} reports few errors.

  \item
    In \mnonstrict{}, the number of agreements jumps up to nearly \pct{20}.
    This is reassuring because type and \FS{} analysis should agree to some extent,
    but the level of agreement is still rather low.

@@

@@

On the meta level, the main takeaway is that the lightweight, counter-based telemetry
in Roblox Studio supports a variety of useful analyses:
\begin{itemize}
  \item
    Counting specific type errors helps to identify trends and to confirm that
    undesirable events rarely happen (e.g.,\code{UnificationTooComplex}).
  \item
    Testing the relationship between two counts can yield insights.
    The gap between \mstrict{} and \FS{} errors in~\cref{f:tefs-compass}
    revealed an issue with \mstrict{} checking.
  \item
    Reporting the overlap between type errors and the current (approximate) edit
    range lets us assess the outcome of edits without revealing source code.
    It was especially useful to have overlaps for two sets of type errors: old and current.
  \item
    Timestamps give useful, low-level insights about the frequency of keystrokes,
    the length of sessions, and actions over time.
    The event ordering that timestamps provide was invaluable; e.g., for learning
    that error deltas bounce up and down rather than rising shaply, then falling~(\cref{f:error-density}).
\end{itemize}
